Microcredit from Delaying Bill Payments

### 2021-01-04 phil_paper/paper_booster_pumps_wjv.tex
empirical strategy looks good

As a pioneer in water infrastructure investments among developing cities, Manila provides a useful context to study the welfare effects of these types of investments.  In 1997, Manila awarded private concession contracts to two private companies to take over for the existing public utility.  Each company is assigned to provide water to their assigned halves of Metro Manila.  Both companies have gradually replaced the decaying pipe infrastructure inherited from the public utility.

To measure the effect of pipe replacements, one of the companies has provided a map of their water pipes including the installation year for each pipe as well as water billing records for their 1.5 million water connections from January, 2008 to June, 2015.  To measure each water connection's exposure to pipe replacement projects, connections are first located within ``small areas'' --- the smallest geographic designations used by the company each containing around 270 connections.  Next, since projects often replace many pipes in the same place at the same time, each small area is assigned a ``pipe replacement year'' according to the year when the greatest total length of tertiary pipes were installed within that small area.\footnote{Tertiary pipes installed in the replacement year account for 85\% of the total tertiary pipe length in small areas.}  Tertiary pipes act as local feeders transporting water from large primary and secondary pipes directly to households.  Therefore, tertiary pipe replacements likely affect service quality within local areas while primary and secondary pipe replacements may have diffuse impacts throughout the pipe network.  By excluding primary and secondary pipe replacements, this approach limits the potential for spillover effects of pipe replacements onto neighboring areas at the cost of potentially underestimating the total welfare gains from these projects.

Billing records measure monthly water prices and usage for each water connection, which may serve multiple households.  In order to link connection-level consumption to household welfare, billing records for each connection are merged to a survey of water connections conducted by an independent evaluator to monitor compliance with the water utility's service obligations.  This connection survey records demographics for the household that owns the connection as well as the number of other households and people that also use the connection.  Household-level water usage is calculated by dividing total connection usage by the number of households sharing the connection.  The survey also includes household demographics for the connection owner, different measures of water service quality, as well as household investments in booster pumps, water filters, and water storage tanks.

The survey randomly interviewed water connection owners covering 15,000 connections in 2008, 24,000 connections in 2010, and 23,000 connections in 2012.  Because a similar sampling design was followed across survey rounds, 13\% of connections were interviewed in two years, and 1.4\% of connections were interviewed in three years while remaining connections were interviewed in only one year.  Since billing records form a monthly-panel, survey responses must be interpolated

Since billing records are

To examine household demographics, billing data are merged at the connection-level to a water connection survey conducted independently to monitor the quality of the utility's service.

\footnote{Households using connections alone tend to be larger and wealthier than households that share connections with other households according to previous research (\cite{wjv}).}

@@

The water connection survey also includes measures

Additional survey data also includes information on demographics and water quality as well as investments!

Water pipes are linked to areas... 
The utility locates water connections within small areas each containing around 270 connections.

Each small-area is assigned a pipe replacement year as the year when the greatest total length of tertiary pipes are installed within the small area.   This pattern is consistent with

  This calculation includes only tertiary pipes (excluding )

 which allows for linking

The pipe map documents the installation year for each length of pipe, which forms the basis for the pipe replacement measure.   The

@@

Section: Data

	- only households connected earlier (explain) (what share are those?!)
	- consumption per HH
	- reweight data for household level?! YES!
	- survey panel and other panel; date representation, how its imputed, etc.
	- Put in a table describing pipe-replacement;  describe staggered approach; predict pipe-replacement?
	- industrial/commercial footnote later!!

Consumption is measured according to average consumption for not-shared households!  What extra assumption is that!!!  NEED ROBUSTNESS!!!

MEASURING AVERAGE PRICES

\section{Descriptives}

Descriptive evidence indicates that fixing pipes improves water pressure, quality, and reliability, which may each affect consumer welfare.  Table~\ref{table:descriptives} tracks water service improvements by comparing average household survey responses before and after pipe replacement.   [summarize the findings]

Consumers also invest in products and behaviors to compensate for low piped water service quality.  These investments help reveal which aspects of piped water service are most valuable to consumers as well as most affected by pipe replacement.

Large investments in booster pumps suggest that households strongly value water pressure.  Booster pumps are both expensive to purchase and operate. In Manila, booster pumps range in price from 1,200 to 15,000 PhP, which represents a large expense given average monthly household incomes of \input{tables/median_inc}PhP.\footnote{Figures come from scraping results from the first page of searching ``booster pumps'' on the popular online marketplace in Manila, \url{https://www.lazada.com.ph/},  yielding 24 entries with horsepower listed.  Average monthly household income is computed for Metro Manila using the 2015 Family Income and Expenditure Survey.}  The average booster pump uses a 0.9 horsepower engine, which generates monthly costs of around 486 PhP at prevailing energy prices.\footnote{A 1 horsepower engine uses around 0.786 Kw per hour.  Table~\ref{table:descriptives} indicates that households use their booster pumps for 2.6 hrs per day, which implies around 78 hours per month.  In 2012, tariffs for the electricity utility in Manila averaged around 8.8 PhP per KwH.}   Before pipe replacement, 40\% of households invest in booster pumps that increase water pressure.  After pipe replacement, the share of households using booster pumps drops to 15\%.  This finding is consistent with booster pumps providing an important substitute to pressure from new water pipes.
% = .9*.786*78*8.8
% \footnote{A 1 horsepower engine uses around 0.786 Kw per hour.  Table~\ref{table:descriptives} indicates that households use their booster pumps for 2.6 hrs per day, which implies around 78 hours per month.  In 2012, tariffs for the electricity utility in Manila averaged around 8.8 PhP per KwH.}
Small investments in water filters combined with frequent purchases of filtered water both before and as well as after pipe replacement indicate that households may not derive large benefits from improvements in piped water quality.  Only 12\% of households use water filters both before and after pipe replacements.  Low filter usage may stem from the fact that less than half of households report drinking from the tap while over 70\% of households report purchasing filtered water from local water-refilling stations.  These behaviors remain constant after pipe replacements.  Taken together, these findings indicate that water quality improvements from pipe replacements may not be primary drivers of changes in household welfare.

@@

Increases in water pressure from pipe replacements may lead households to increase their monthly water consumption.  Rapid water flow allows households to complete a greater number of water-using activities like cleaning and bathing in the same amount of time.  Greater water pressure may also allow households to engage in new activities that require minimum pressure levels such as showering.\footnote{By contrast, water quality improvements may have an ambiguous effect on water consumption.  On one hand, cleaner water may induce households to use more tap water for cooking and cleaning.  On the other hand, cleaner water may increase the productivity of cleaning and bathing, which may lead households to use less piped water.}  Figure~\ref{figure:pipecons} plots average monthly water usage per household in the 4 years before and 6 after pipe replacement.  Usage increases from an average of \input{tables/c_pre}m3 before pipe replacement to \input{tables/c_post}m3 after pipe replacement, which represents an \input{tables/c_diff_per}\unskip\% increase.  The increase in usage occurs abruptly at the year of pipe replacement and remains at roughly the same level in the following 6 years.  This sustained increase in consumption suggests that pipe replacement may provide sustained impacts on household welfare.

The absence of strong pre-trends in usage suggests that replacement projects are not targeted to areas with particular trends in local water demand.  Instead, this pattern is consistent with the company's stated goal of sequentially replacing old pipes according to engineering specifications.  Table~\ref{table:descriptives} further supports this theory by indicating few demographic differences between households that receive pipe replacement projects (columns (1) and (2)) and all households (column (3)).  Also, demographic characteristics appear relatively similar before and after pipe replacement, which suggests that pipe replacements were not coupled with other policies that may have also affected demand for water.

@@

Mapping these increases in consumption into household welfare requires measuring how households trade off water usage and price.  Prices are determined by the government regulator who imposes an increasing tariff schedule according to monthly usage, which is standard among public utilities (\cite{hoque2013state}).  Despite steep increases in marginal price at specific levels of usage, households do not appear to be sensitive to these price changes because households are not observed adjusting their consumption strategically to avoid higher prices (cite appendix).  The regulator also increases prices gradually over time to ensure that the company continues to cover operating costs.  Since households also gradually increase their average usage over this interval, it is unclear whether households are responsive to these price increases (cite appendix).

[ RE-WRITE WITH NEW DEFINITION! ]
The government regulator also gives the water company discretion to assign households to a high-price tariff schedule if they have any business activity at their residence, which provides a novel source of price variation in the context of public utilities.  In the vast majority of cases, the high price tariff is applied to households that operate small food stands (or ``Sari-Sari'' stores).  For a household with average monthly usage, the high-price tariff results in an average price of \input{tables/p_s}PhP/m3 while the regular tariff results in an average price of \input{tables/p_r}PhP/m3 (cite appendix for tariff).  The water company periodically visits consumers and updates prices according to the activities observed at each consumer's residence.  In some cases, consumers request price changes, which prompts the company to investigate the household and determine the appropriate price.  Table~\ref{table:pricechangestatistics} provides average characteristics of households that are always observed with the regular price (in column (1)), that are always observed with the high price (in column (2)), and that are observed changing prices during the sample (in column (3)).  While the 2,442 households that experience price changes use more water than other households, they share similar demographic characteristics, which suggests that they may also share similar price-sensitivities.

Figure~\ref{figure:usagepricechanges} plots average usage 4 years before and after households are switched from the regular price to the high price (in red) as well as before and after households are switched from the high price to the regular price (in blue).  Before the price change, average usage for both groups follows relatively constant trends.  The price change is associated with a usage jump for households switched to the regular price and a corresponding usage slide for households switched to the high price.  The changes in consumption appear relatively persistent up to 4 four years after the price changes.  These patterns indicate that household water usage is sensitive to large, discrete price changes.

@@

\begin{figure}
\begin{center}
\caption{Usage and Price Changes}\label{figure:usagepricechanges}
\includegraphics[scale=1]{tables/r_to_s_graph.pdf}
\end{center}
\end{figure}

@@

Each month, household utility takes the following form
\begin{align}
\label{eq:utility}
U\,=\,\frac{1}{\alpha} \, \Big[ \,  Q(B,R) \,  w  \, -\, \frac{1}{2}(w - \gamma)^2 \, \Big] \, + \, x 
\end{align}
where $w$ is water consumption and $x$ represents a bundle of all other goods consumed by the household.  Water service quality given by the function, $Q(B,R)$, depends on booster pump use where $B$ equals 1 if the household chooses to use a booster pump (and zero otherwise) and as well as pipe replacements where $R$ equals 1 after pipe replacement (and zero otherwise).  $Q(B,R)$ is assumed to be differentiable in $R$.  Water service quality enters multiplicatively with water consumption, which nests the assumption that each unit of water consumption is affected equally and positively by water service quality.  This assumption also excludes the possibility that improved water service quality affects household utility in ways that are not directly proportional to household water consumption such as by increasing housing values.  In this case, the approach would underestimate the welfare benefits of improved service quality.  $\alpha$ reflects price sensitivity, and $\gamma$ reflects the satiation amount of water usage.  This approach assumes that preferences for water are quasi-linear, which implies that water consumption does not depend on household income [ Footnote to heterogeneity table ].

@@

Households maximize utility subject to the following budget constraint
\begin{align}
\label{eq:bc}
p\,w \,+\, x \, +  B \, F \, = Y
\end{align}
where $p$ is the average price of water while the price of all other goods, $x$, is normalized to 1.  This approach includes the simplifying assumption that households respond to a single, average water price $p$ despite facing marginal prices that increase with monthly consumption.  This assumption is consistent with descriptive evidence that households appear unresponsive to marginal prices (cite appendix).   $F$ is total cost of renting and using a booster pump each month.  This approach assumes that Manila has a competitive market for renting/selling booster pumps that is unaffected by improvements in piped water service quality.\footnote{This assumption is consistent with local service quality improvements and a city-wide market for booster pumps.}  $Y$ is monthly household income.

@@

Maximizing household utility in equation $(\ref{eq:utility})$ subject to the budget constraint in equation $(\ref{eq:bc})$ for a given choice of booster pump usage, $B$, yields the following expression for monthly water demand
\begin{align}
\label{eq:demand}
w^{*} \, = \, \gamma \, - \, \alpha p +  Q(B,R)
\end{align}
Water demand depends linearly on the satiation preference, $\gamma$, price, and service quality.

Given optimal water demand, the indirect utility function at the optimal booster pump choice, $B^{*}$, is given by 
\begin{align}
V\,=\,\frac{\alpha p^2}{2} - \gamma p + \frac{Q(B^{*},R)^{2}}{2\alpha} - p Q(B^{*},R) + \frac{\gamma Q(B^{*},R)}{\alpha} + y - B^{*} F
\end{align}
The derivative of $V$ with respect to pipe replacements, $R$, reflects the change in consumer welfare associated with pipe replacements.  This derivative takes the following form after substituting terms for $w^{*}$
\begin{align}
\label{eq:dvdr}
\frac{dV}{dR}\,=\,\frac{w^{*}}{\alpha} \frac{dQ}{dR} - F \frac{dB^{*}}{dR}
\end{align}
This expression summarizes the consumer welfare effects of pipe replacements in terms of (1) the marginal effect on service quality weighted by water consumption and price sensitivity as well as (2) the marginal effect on booster pump usage weighted by the cost of booster pumps.  This approach assumes that welfare effects are proportional to consumption, $w^{*}$, implying that service quality improvements affect all units of water consumption.  Welfare effects are decreasing in price sensitivity, $\alpha$, which captures the intuition that households with many substitute water sources (ie. high $\alpha$) may benefit less from improvements in service quality.  According to this approach, welfare changes do not require measuring fixed preferences, $\gamma$, separately from levels of service quality, $Q(B^{*},R)$.  Therefore, this approach is robust to different patterns of selection into booster pump use based on different levels of service quality or fixed preferences for water consumption.

@@

The estimation strategy recovers household price sensitivity as well as the changes in service quality and booster pump use in response to pipe replacement projects. These estimated quantities along with observed water use and booster pump costs summarize the change in consumer welfare from pipe replacement projects as given by equation (\ref{eq:dvdr}).  The estimating equation takes the following form
\begin{align}
\label{eq:esteq}
y_{itl} \,=\, \beta_1 \, \text{After Pipe Replacement}_{tl} \,+\, \beta_2 \, p_{it} \, + \, \theta_t \, + \, \lambda_i \, + \epsilon_{itl}
\end{align}
where the outcome, $y_{itl}$, is either monthly water consumption or booster pump use.  Outcomes are measured for household, $i$, in small area, $l$, at calendar month, $t$. After Pipe $\text{Replacement}_{tl} $ takes a value of one for months after pipe replacement and zero otherwise.  Since the data identify only the year that pipes were replaced, all months from the replacement year onwards are considered after pipe replacement. $p_{it}$ measures the average price faced by each household in each month. $\theta_t$ includes calendar month fixed effects while $\lambda_i$ includes household fixed effects.

When the outcome is monthly water consumption, the estimating equation maps directly onto the equation for optimal household consumption given by equation ($\ref{eq:demand}$).  $\beta_1$ captures the effect of pipe replacement on service quality, $\frac{dQ}{dR}$.  Likewise, $\beta_2$ captures the negative of household price sensitivity, $\alpha$.  Remaining fixed preferences for water and levels of service quality are absorbed in household and month fixed effects, $\lambda_i$ and $\theta_t$.  These fixed effects ensure that the effect of pipe replacements is identified from variation in water consumption for the same household before and after pipes are fixed while accounting for any temporal consumption patterns common to all households.  Likewise, price sensitivity is estimated from variation in price for the same household switching between regular and high prices.  Calendar month fixed effects absorb price variation affecting all households equally over time as the company updates water tariffs.

Identifying $\beta_1$ and $\beta_2$ requires assuming that pipe replacements and household-specific price changes are not correlated with any other factors that may also affect water demand at the same time.  One concern may be that pipe improvements are paired with other infrastructure investments that drive increases in local income, population, and water demand.  Alternatively, the water company may strategically target pipe improvements to either boost areas with declining demand or accelerate areas with growing demand.  Figure~\ref{figure:pipecons} traces a sharp break in average usage at the year of pipe replacement with little evidence of strong trends before or after replacement projects.  These patterns are consistent with company reports that pipe replacements were not explicitly linked to other infrastructure projects and were instead planned primarily to minimize engineering costs.

Another concern is that households may increase their water demand as they are switched to the high price because they often open roadside food-stands.  Conversely, households that close their roadside food-stands may decrease their consumption at the same time as they are switched to the low price.  Both cases suggest that the empirical approach would underestimate household price sensitivities.  Figure~\ref{figure:usagepricechanges} provides little evidence of strong pre-trends leading up to price changes.  While this evidence is suggestive that these events are uncorrelated with long-term trends in water demand, this approach is unable to exclude the possibility that other factors may influence water demand at the same time as these events.

>>>>

As a pioneer in water infrastructure investments among developing cities, Manila provides a useful context to study the welfare effects of these types of pipe replacements.  Before 1997, a single government utility provided water to Manila resulting in low levels of service quality (\cite{dumol2000manila}).  In 1997, Manila awarded private concession contracts to two private companies to take over for the public utility.  Each company was assigned to provide water to their assigned halves of Metro Manila.  The two companies are regulated by a government agency who sets the water tariff in order to ensure that the companies are able to cover their costs while earning a fair rate of return on their assets.  The regulator also reserves the right to disallow costs that it views as inessential to water provision.  Privatization led to large improvements in water access such that by 2015, piped water usage was almost universal.\footnote{Less than 2\% of households report using well water, water peddlers, or other alternative water sources for cooking according to the 2015 Census of Population and Housing.}  This paper analyzes data provided by one of the companies.

In 2008 in agreement with the government regulator, the water company started aggressively replacing the decaying pipe infrastructure that it inherited from the public utility.  Guided by hydrological concerns, the company separated the service area into smaller sections and replaced most pipes in each section on a sequential basis.  The existing pipes were mostly installed between 1980 and 1990 with an average year of 1986.\footnote{DISCUSS APPENDIX ABOUT PIPE AGE.}  Between 2008 and 2015, the water company replaced over 57\% of its nearly 6,000 km of pipes.

To measure the effect of pipe replacements, the company provided a map of their water pipes including the installation year for each pipe as well as water billing records for their 1.5 million water connections from January, 2008 to June, 2015.\footnote{Data is missing for the month of June, 2014.}  To measure each water connection's exposure to pipe replacement projects, connections are first located within ``small areas'' --- the smallest geographic designations used by the company with 2,976 total areas each containing around 270 connections.  Next, since projects often replace many pipes in the same place at the same time, each small area is assigned a ``pipe replacement year'' according to the year when at least 80\% of the total length of tertiary pipes were installed within that small area.  576 small areas are observed before and after their ``pipe replacement year.''  Tertiary pipes act as local feeders transporting water from large primary and secondary pipes directly to households.  Therefore, tertiary pipe replacements likely affect service quality within local areas while primary and secondary pipe replacements may have diffuse impacts throughout the pipe network.  By excluding primary and secondary pipe replacements, this approach limits the potential for spillover effects of pipe replacements onto neighboring areas at the cost of potentially underestimating the total welfare gains from these projects.  % \footnote{Tertiary pipes installed in the replacement year account for 85\% of the total tertiary pipe length in small areas.}

Billing records measure monthly water prices and usage for each water connection, which may serve multiple households.  In order to link connection-level consumption to household welfare, billing records for each connection are merged to a survey of residential water connections conducted by an independent evaluator to monitor compliance with the water utility's service obligations.\footnote{The billing records also cover commercial and industrial accounts.}  This connection survey records demographics for the household that owns the connection as well as the number of other households and people that also use the connection.  The survey also includes household demographics for the connection owner, different measures of water service quality, as well as household investments in booster pumps, water filters, and water storage tanks.

The main outcome of interest is household-level water usage, which is calculated by dividing total connection usage by the number of households sharing the connection.  Outlier months with over 200 m3 of consumption are excluded from the analysis.  All empirical results are weighted at the household-level to ensure that results are representative of the population of households.

The survey randomly interviewed water connection owners covering 15,000 connections in 2008, 24,000 connections in 2010, and 23,000 connections in 2012.  Because a similar sampling design was followed across survey rounds, 13\% of connections were interviewed in two years, and 1.4\% of connections were interviewed in three years while remaining connections were interviewed in only one year.  The connection survey is merged to the billing records by the connection id number and the interview month.  For billing-record months without corresponding interviews, survey responses are interpolated with responses from the most recent interview.

Since households face an increasing block water tariff, the average water price faced by households is calculated as the average  amount paid per cubic meter of water usage for each calendar month and tariff category after dropping outlier values below 5 PhP/m3 and above 100 PhP/m3.\footnote{Households either face a standard tariff or a high-price tariff depending on whether they are observed with business activities at their residence.  See Section~\ref{section:descriptives} for more discussion.}  Outliers are often due to billing credits/errors.  To measure savings from pipe replacements, the water company also provided records of project costs as well as the quantity of water supplied, which often exceeds the quantity of water billed.  These attributes are measured according to ``district metering areas,'' which are around three times larger than small areas and are delineated based on hydrological considerations in order to monitor water loss.

% DISCUSS NRW AND COMM ACCOUNTS HERE!!

% MENTION LATER!: The billing records also measure consumption for commercial and industrial connections, which are used to measure

% MEASURING AVERAGE PRICES

% To examine household demographics, billing data are merged at the connection-level to a water connection survey conducted independently to monitor the quality of the utility's service.

% \footnote{Households using connections alone tend to be larger and wealthier than households that share connections with other households according to previous research (\cite{wjv}).}

@@

@@

% Section: Data

	% - only households connected earlier (explain) (what share are those?!)
	% - consumption per HH
	% - reweight data for household level?! YES!
	% - survey panel and other panel; date representation, how its imputed, etc.
	% - Put in a table describing pipe-replacement; describe staggered approach; predict pipe-replacement?
	% - industrial/commercial footnote later!!

% Consumption is measured according to average consumption for not-shared households!  What extra assumption is that!!!  NEED ROBUSTNESS!!!

\section{Descriptives}\label{section:descriptives}

Descriptive evidence indicates that fixing pipes improves water pressure, quality, and reliability, which may each affect consumer welfare.  Table~\ref{table:descriptives} tracks water service improvements by comparing average household survey responses before and after pipe replacement for areas that are observed both before and after.\footnote{Areas that may have experienced pipe replacements before the start of the sample are included in the ``All'' category.}  The share of households reporting strong water pressure during peak evening usage from 6pm to 12am increases from 25\% to 58\% after pipe replacements.  Likewise, the share reporting no pressure over the same interval decreases from 29\% to 4\%.  Water interruptions decline slightly from 2.28 to 1.93 in the last three months, which is consistent with the water company facing periodic droughts.  Water quality shows large improvements with the share of respondents reporting discolored, unusual tasting, or contaminated water dropping by at least 5 percentage points for each measure.

Households often invest in products and behaviors to compensate for low piped water service quality.  These investments help reveal which aspects of piped water service are most valuable to consumers as well as most affected by pipe replacement.

Large investments in booster pumps suggest that households strongly value water pressure and reliability.  Booster pumps are both expensive to purchase and operate. In Manila, booster pumps range in price from 1,200 to 15,000 PhP, which represents a large expense given average monthly household incomes of \input{tables/median_inc}PhP.\footnote{Figures come from scraping results from the first page of searching ``booster pumps'' on the popular online marketplace in Manila, \url{https://www.lazada.com.ph/},  yielding 24 entries with horsepower listed.  Average monthly household income is computed for Metro Manila using the 2015 Family Income and Expenditure Survey.}  The average booster pump uses a 0.9 horsepower engine, which generates monthly costs of around 486 PhP at prevailing energy prices.\footnote{A 1 horsepower engine uses around 0.786 Kw per hour.  Table~\ref{table:descriptives} indicates that households use their booster pumps for 2.6 hrs per day, which implies around 78 hours per month.  In 2012, tariffs for the electricity utility in Manila averaged around 8.8 PhP per KwH.}   Before pipe replacement, 40\% of households invest in booster pumps that increase water pressure.  After pipe replacement, the share of households using booster pumps drops to 15\%.  This finding is consistent with booster pumps providing an important substitute to pressure from new water pipes.
% = .9*.786*78*8.8
% \footnote{A 1 horsepower engine uses around 0.786 Kw per hour.  Table~\ref{table:descriptives} indicates that households use their booster pumps for 2.6 hrs per day, which implies around 78 hours per month.  In 2012, tariffs for the electricity utility in Manila averaged around 8.8 PhP per KwH.}

Small investments in water filters combined with frequent purchases of filtered water both before and as well as after pipe replacement indicate that households may derive fewer benefits from improvements in piped water quality.  Only 12\% of households use water filters both before and after pipe replacements.  Low filter usage may stem from the fact that less than half of households report drinking from the tap while over 70\% of households report purchasing filtered water from local water-refilling stations.  These behaviors remain constant after pipe replacements.  Taken together, these findings indicate that water quality improvements from pipe replacements may not be primary drivers of changes in household welfare.

@@

Increases in water pressure from pipe replacements may lead households to increase their monthly water consumption.  Rapid water flow allows households to complete a greater number of water-using activities like cleaning and bathing in the same amount of time.  Greater water pressure may also allow households to engage in new activities that require minimum pressure levels such as showering.  Cleaner water may induce households to use more tap water for cooking and cleaning.  Figure~\ref{figure:pipecons} plots average monthly water usage per household in the 4 years before and 6 after pipe replacement.  Usage increases from an average of \input{tables/c_pre}m3 before pipe replacement to \input{tables/c_post}m3 after pipe replacement, which represents an \input{tables/c_diff_per}\unskip\% increase.  The increase in usage occurs abruptly at the year of pipe replacement and remains at roughly the same level in the following 6 years.  This sustained increase in consumption suggests that pipe replacements may provide long-term impacts on household welfare.

The absence of strong pre-trends in usage suggests that replacement projects are not targeted to areas with particular trends in local water demand.  Instead, this pattern is consistent with the company's stated goal of sequentially replacing old pipes according to hydrological specifications.  Table~\ref{table:descriptives} further supports this theory by indicating few demographic differences between households that receive pipe replacement projects (columns (1) and (2)) and all households (column (3)).  Also, demographic characteristics appear relatively similar before and after pipe replacement, which suggests that pipe replacements were not coupled with other policies that may have also affected demand for water.

@@

Mapping these increases in consumption into household welfare requires measuring how households trade off water usage and price.  Prices are determined by the government regulator who imposes an increasing tariff schedule according to monthly usage, which is standard among public utilities (\cite{hoque2013state}).  Despite steep increases in marginal price at specific levels of usage, households do not appear to be sensitive to these price changes because households are not observed adjusting their consumption strategically to avoid higher prices.\footnote{Appendix Figure~\ref{figure:tariffschedule} includes the average tariff schedule for users facing low and high prices.  Appendix Figure~\ref{figure:conshist} includes a histogram of usage per connection and finds little evidence of significant bunching at the tariff kink points at 10, 20, and 40 m3.}  The regulator also increases prices gradually over time to ensure that the company continues to cover operating costs.  Since households also gradually increase their average usage over this interval, it is unclear whether households are responsive to these price increases.\footnote{Appendix Figure~\ref{figure:pricetimeseries} plots average prices and usage per household over time.}

% [ RE-WRITE WITH NEW DEFINITION! ]
The empirical exercise analyzes an event-study where the same households are exposed to very different price levels over time.  The government regulator gives the water company discretion to assign households to a high-price tariff schedule if they have any business activity at their residence, which provides a novel source of price variation in the context of public utilities.  In the vast majority of cases, the high price tariff is applied to households that operate small food stands (or ``Sari-Sari'' stores).  For a household with average monthly usage, the high-price tariff results in an average price of \input{tables/p_s}PhP/m3 while the regular tariff results in an average price of \input{tables/p_r}PhP/m3.\footnote{Appendix~\ref{figure:tariffschedule} plots the tariff.}  The water company occasionally visits consumers and assigns consumers to the high-price tariff if business activity is observed at the residence.\footnote{In some cases, consumers request price decreases when they close their businesses.  These price-decrease events are not the focus of the analysis because consumers directly anticipate these events and may otherwise have adjusted their water usage accordingly.}  Table~\ref{table:pricechangestatistics} provides average characteristics of households that are observed with the regular price (in column (1)), that are observed with the high price (in column (2)), and that are observed increasing prices during the sample (in column (3)).  While the 724 households that experience price increases use more water than other households, they share similar demographic characteristics, which suggests that they may also share similar price-sensitivities.

Figure~\ref{figure:usagepricechanges} plots average usage and marginal prices 4 years before and after households are first switched from the regular price to the high price.\footnote{In very few cases, households are later switched back to the regular price, and then switched again to the high price.}  Before the price change, average usage increases likely due to some households using additional water for their small-businesses.  The price change at month 0 is associated with a sharp drop in usage and increase in average prices that persist for at least 24 months.  These patterns indicate that household water usage is sensitive to large, discrete price changes over time.

@@

\begin{figure}
\begin{center}
\caption{Usage and Price Changes}\label{figure:usagepricechanges}
\includegraphics[scale=1]{tables/r_to_s_only_graph.pdf}
\end{center}
\end{figure}

@@

Each month, household utility takes the following form
\begin{align}
\label{eq:utility}
U\,=\,\frac{1}{\alpha} \, \Big[ \,  Q(B,R) \,  w  \, -\, \frac{1}{2}(w - \gamma)^2 \, \Big] \, + \, x 
\end{align}
where $w$ is water usage and $x$ represents a bundle of all other goods consumed by the household.  Water service quality given by the function, $Q(B,R)$, depends on booster pump use where $B$ equals 1 if the household chooses to use a booster pump (and zero otherwise) and as well as pipe replacements where $R$ equals 1 after pipe replacement (and zero otherwise).  $Q(B,R)$ is assumed to be differentiable in $R$.  Water service quality enters multiplicatively with water usage, which nests the assumption that each unit of water usage is affected equally and positively by water service quality.  This assumption also excludes the possibility that improved water service quality affects household utility in ways that are not directly proportional to household water usage such as by increasing housing values.  In this case, the approach would underestimate the welfare benefits of improved service quality.  $\alpha$ reflects price sensitivity, and $\gamma$ reflects the satiation amount of water usage.  This approach assumes that preferences for water are quasi-linear, which implies that water usage does not depend on household income [ Footnote to heterogeneity table ].

@@

Households maximize utility subject to the following budget constraint
\begin{align}
\label{eq:bc}
p\,w \,+\, x \, +  B \, F \, = Y
\end{align}
where $p$ is the average price of water while the price of all other goods, $x$, is normalized to 1.  This approach includes the simplifying assumption that households respond to a single, average water price $p$ despite facing marginal prices that increase with monthly usage.  This assumption is consistent with descriptive evidence that households appear unresponsive to marginal prices.\footnote{Appendix Figure~\ref{figure:conshist} includes a histogram of usage per connection and finds little evidence of significant bunching at the tariff kink points at 10, 20, and 40 m3.}   $F$ is total cost of renting and using a booster pump each month.  This approach assumes that Manila has a competitive market for renting and selling booster pumps that is unaffected by improvements in piped water service quality.\footnote{This assumption is consistent with local service quality improvements and a city-wide market for booster pumps.}  $Y$ is monthly household income.

@@

Maximizing household utility in equation $(\ref{eq:utility})$ subject to the budget constraint in equation $(\ref{eq:bc})$ for a given choice of booster pump usage, $B$, yields the following expression for monthly water demand
\begin{align}
\label{eq:demand}
w(B) \, = \, \gamma \, - \, \alpha p +  Q(B,R)
\end{align}
Water demand depends linearly on the satiation preference, $\gamma$, price, and service quality.

Given optimal water demand, the indirect utility function at a given booster pump choice, $B$, is given by 
\begin{align}
V(B)\,=\,\frac{\alpha p^2}{2} - \gamma p + \frac{Q(B,R)^{2}}{2\alpha} - p Q(B,R) + \frac{\gamma Q(B,R)}{\alpha} + y - B F
\end{align}
The derivative of $V$ with respect to pipe replacements, $R$, reflects the change in consumer welfare associated with pipe replacements.  This derivative takes the following form after substituting terms for $w(B)$
\begin{align}
\label{eq:dvdr}
\frac{dV(B)}{dR}\,=\,\frac{w(B)}{\alpha} \frac{dQ}{dR} - F \frac{dB}{dR}
\end{align}
This expression summarizes the consumer welfare effects of pipe replacements in terms of (1) the marginal effect on service quality weighted by water usage and price sensitivity as well as (2) the marginal effect on booster pump usage weighted by the cost of booster pumps.  This approach assumes that welfare effects are proportional to usage, $w^{*}$, implying that service quality improvements affect all units of water usage.  Welfare effects are decreasing in price sensitivity, $\alpha$, which captures the intuition that households with many substitute water sources (ie. high $\alpha$) may benefit less from improvements in service quality.  According to this approach, welfare changes do not require measuring fixed preferences, $\gamma$, separately from levels of service quality, $Q(B,R)$.  Therefore, this approach is robust to different patterns of selection into booster pump use based on different levels of service quality or fixed preferences for water usage.

@@

The empirical strategy estimates household price sensitivity as well as changes in water usage and booster pump use in response to pipe replacement projects. These estimated quantities inform the effect on consumer welfare of pipe replacement projects given by equation (\ref{eq:dvdr}).  Using an instrumental variables approach, the empirical strategy isolates price variation from connections being switched into high price tariffs.  The first stage equation takes the following form
\begin{align}
\label{eq:1ststage}
p_{it} &= \phi_1 \text{Post Pipe}_{tl}  +  \phi_2 \text{Post Price}_{it} +  \phi_3\text{Pre Trend}_{it} +\phi_4 \, \text{Post Trend}_{it} +  \phi_t  +  \phi_{il} + \epsilon_{itl} 
\end{align}
where $i$ indexes the water connection, $t$ indexes calendar month, and $l$ indexes small area.  $p_{it}$ is the average price per cubic meter of monthly water use.  $\text{Post Pipe}_{tl}$ is an indicator for months during and after January of the pipe replacement year.  $\text{Post Price}_{it}$ is an indicator for months after a connection is first switched into a high-price tariff.\footnote{In rare instances, connections may be switched into high-price tariffs more than once during the study period.  The instrumental variables design ensures that the estimation uses price variation only from the first price increase, which is more likely to be unexpected by households than later price increases.}  $\text{Post Price}_{it}$ is the instrumental variable that is excluded in the second stage.  To account for trends in water usage around the time of the price change, $\text{Pre Trend}_{it}$ counts the months leading up to the first price increase and takes a value of zero otherwise.  Likewise, $\text{Post Trend}_{it}$ counts months after the first price increase and takes a value of zero otherwise.  $\phi_t$ are calendar month fixed effects.  $\phi_{il}$ are connection fixed effects when the second stage has water usage as an outcome.  $\phi_{il}$ are small-area fixed effects when the second stage has booster pump use as an outcome because booster pump use is only observed in months where the connection survey was fielded.

Given predicted prices from the first stage equation, the second stage equations take the following form
\begin{align}
w_{itl} &= \gamma_1  \text{Post Pipe}_{tl} + \gamma_2  \hat{p}_{it} + \gamma_3\text{Pre Trend}_{it} +\gamma_4\text{Post Trend}_{it} +    \gamma_t +  \gamma_i +  \varepsilon_{itl}  \label{eq:2ndstage1} \\
b_{itl} &= \beta_1  \text{Post Pipe}_{tl} + \beta_2  \hat{p}_{it} + \beta_3\text{Pre Trend}_{it} +\beta_4\text{Post Trend}_{it} +    \beta_t +  \beta_l +  \varepsilon_{itl} \label{eq:2ndstage2}
\end{align}
where $w_{itl}$ is water usage per household in cubic meters per month and $b_{itl}$ is an indicator for connection booster pump usage.

The coefficients of interest for the impacts of pipe replacement are given by $\gamma_1$ and $\beta_1$ for water and booster pump use respectively.  After controlling for calendar month and connection fixed effects, these coefficients are identified from differential changes in outcomes for areas that experienced pipe replacements compared to areas that did not over the same interval.  These coefficients reflect the causal effect of pipe replacements under the assumption that no other factors changed at the same time that would otherwise affect the outcomes.  This assumption is supported by qualitative evidence that pipe replacements are often driven by hydrological, engineering concerns rather than strategic responses to changes in local water demand or other consumer characteristics.  Supporting this assumption, Figure~\ref{figure:pipecons} traces a sharp break in average usage at the year of pipe replacement with little evidence of strong trends before or after replacement projects.
